---
title: "Project Part 1"
author: "Het Patel (7972424)"
date: "9 Nov 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
nba <- read.csv("nbaplayers.csv")
```

DESCREPTION: This is dataset of 50 NBA players from 2022-23 regular season. This dataset has total of 539 players but because of the project I have taken the first 50 players from it. It has the following columns:

1)  **Rank**: The standing of the player
2)  **PLAYER**: Name of the player
3)  **TEAM**: Team in which player play
4)  **GP**: Games played in the season
5)  **PTS**: Average points per game in that season
6)  **FGA**: Field goals attempted per game in that season
7)  **FTA**: Free throws attemplted per game in that season

Here's the URL where I found my dataset: [dataset link](https://www.nba.com/stats/players/traditional?PerMode=Totals&sort=PTS&dir=-1&Season=2022-23)

My response variable is PTS (average points per game)

My two explanatory variables are FGA (average field goals attempted) and FTA (Free throws attemplted per game)

```{r}
nba2.lm <- lm(PTS ~ FGA, data = nba)
summary(nba2.lm)

# R^2 = 0.5554, Adj_r^2 = 0.5461


nba3.lm <- lm(PTS ~ FTA, data = nba)
summary(nba3.lm)

#R^2 = 0.595, Adj_r^2 = 0.5866
```

For $Y\sim X1$, $r^2=0.5554$ and $R^2_{adj}=0.5461$ For $Y\sim X2$, $r^2=0.595$ and $R^2_{adj}=0.5866$

```{r}
nba1.lm <- lm(PTS ~ FGA + FTA, data = nba)
summary(nba1.lm)

# Adj_r^2 = 0.8578
```

For $Y\sim X1 + X2$, $R^2_{adj}=0.8578$
