---
title: "Predicting Avg. Points per game of NBA Players - Intro Draft (Project Part 2)"
author: "Het Patel (7972424)"
date: "Nov. 22 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing required libraries
library(kableExtra) # for kable()
```
# Question

## Response Variable:
The response variable that I would like to model is PTS, i.e., average Points per game of a NBA player in 2022-23 regular season.

## Explanatory Variable:
Variable FGA, i.e., average field goals attempted per game and FTA, i.e., average field throws attempted per game for a NBA player in 2022-23 regular season.

The reason why I choose these two variables as my explanatory is because more the number of field and free throws attempted by a player, more the chances that a player will score more points.

# Dataset

```{r}
nba <- read.csv("nbaplayers.csv")
kable(nba, format = "markdown")
```

## Dataset Citation

National Basketball Association.(2023). NBA Advance Player Stats(2022-23 Season)[Data set]. NBA Media Ventures. [https://www.nba.com/stats/players/traditional?PerMode=Totals&sort=PTS&dir=-1&Season=2022-23](https://www.nba.com/stats/players/traditional?PerMode=Totals&sort=PTS&dir=-1&Season=2022-23)

\newpage

## Variables

The dataset contains total of 7 variables and the details for each is as follows:

1) The **Rank** variable is the standing of player in order of average points scored in 2022-23 regular season.

2) The **PLAYER** variable is the name of NBA player.

3) The **TEAM** variable is the name of team in which player played the 2022-23 regular season.

4) The **GP** variable is the number of games played by the player in the 2022-23 regular season.

5) The **PTS** variable is the average number of points scored per game by a player in 2022-23 regular season.

6) The **FGA** variable is the average number of field throws attempted per game by a player in 2022-23 regular season.

7) The **FTA** variable is the average number of free throws attempted per game by a player in 2022-23 regular season.

## Scatterplots

```{r}
plot(nba$FGA, nba$PTS, main = "Field throws attempted (avg) vs Points scored (avg)",
     xlab = "FGA", ylab = "PTS")
```
\newpage
```{r}
rsqr1 <- cor(nba$FGA, nba$PTS)^2
rsqr1
```

The $r^2_{Y \sim X_{1}}$ = 0.555407, which shows that they appear to be in relationship.

```{r}
plot(nba$FGA, nba$PTS, main = "Free throws attempted (avg) vs Points scored (avg)",
     xlab = "FTA", ylab = "PTS")
rsqr2 <- cor(nba$FTA, nba$PTS)^2
rsqr2
```

The $r^2_{Y \sim X_{2}}$ = 0.5949966, which shows that they are correlated with each other.